\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-05-14 paper.qmd
Plan for writing

The 2D hexagon grid is defined by the number of bins in each direction $(b_1, b_2)$, as given by the centroids of each hexagon $(c_1, c_2)$ and the lower left position where the grid starts at $(s_1, s_2)$, which correspond to the lowest left centroid. The values of $s_i$ need to be below their respective minimum variable values, and could be a full bin lower, to allow a buffer ($q$) corresponding to a full hexagon width ($a_1$) and height ($a_2$) around the data. The values of $b_i$ are variables to be computed that define the reduction in size of the data ($n$ to $m$).

@@

@@

<!--We deliberately separate out the creation of the hexagonal grid from the mapping of points on the grid.-->
#### Obtaining bin representations

Let $h : \mathbb{R}^{n\times k} \rightarrow \mathbb{R}^{m\times k}$ be the function that maps a point in \kD{} space to it's bin representation ($C^{(k)}$). This is the model points in \kD{} space.

@@

#### Generating edges

Delaunay triangulation on bin representations generates edges by specifying which bin representations should be connected to generate the model in \kD{} space.

@@

The final step of the algorithm is _lifting_ the \kD{} model back to \pD{}, completing the *model-in-the-data-space* picture. For this, define the set $H^k$ to be the set of points that belong to centroid $k$. That is, $H^k = \{ y \; | \; g(y) = (h_x^k, h_y^k)\}$. We use the \pD{} Euclidean mean of the points in $H^k$ to map the centroid $(h_x^k, h_y^k)$ to a point in \pD{}. Let the $i$-th component of the \pD{} mean be

@@

### Measuring the fit

### What can be learned

- Overview: Generate a form that maps the model, that is the interpoint distances. What is the model?
- Notation
- Create a representation of the model 
    - using hex-binning in 2D, 
    - parameters, 
    - tuning, 
    - pre-processing
- How does this map to the representation in high-d
    - Centroids, 
    - Edges
- Measuring fit
    - Fitted values
    - Error calculation
- What is learned about simulated examples
    - Interesting organisation of points in UMAP
    -

>>>>

The 2D hexagon grid is defined by the number of bins in each direction $(b_1, b_2)$, giving total number of bins as $b = b_1 \times b_2$, and hexagon id, $h = 1, \dots, b$. Each hexagon, $H_h$ is uniquely described by centroid, $ C_{h}^{(2)} = (c_{h1}, c_{h2})$. The lower left position where the grid starts at $(s_1, s_2)$, which correspond to the lowest left centroid. The values of $s_i$ need to be below their respective minimum variable values, and could be a full bin lower, to allow a buffer ($q$) corresponding to a full hexagon width ($a_1$) and height ($a_2$) around the data. The values of $b_i$ are variables to be computed that define the reduction in size of the data ($n$ to $m$).

@@

$\{ i \in H_h, h = 1, \dots, b, \text{ and } i = 1, \dots, n\}$

@@

<!--We deliberately separate out the creation of the hexagonal grid from the mapping of points on the grid.-->
#### Summarization of the data in 2D

Let $h : \mathbb{R}^{n\times k} \rightarrow \mathbb{R}^{m\times k}$ be the function that maps a point in \kD{} space to it's bin centroid ($C^{(k)}$). This is the model points in \kD{} space.

@@

#### Generating edges to indicate neighbours and removing long edges

Delaunay triangulation is used to connect neighbouring centroids, which is needed to preserve neighbourhood information when the model is lifted into \pD{}.

@@

The last step is to lift the \kD{} model into \pD{} by computing \pD{} vectors that represent bin centroids. Define the set $H^k$ to be the set of points that belong to bin $k$. We use the \pD{} Euclidean mean of the points in $H^k$ to map the centroid $(h_x^k, h_y^k)$ to a point in \pD{}. Let the $i$-th component of the \pD{} mean be

@@

<!--langevitour with pD model for S-curve-->

### Measuring the fit
 <!-- Fitted values,  Error calculation-->

### Tuning

<!-- removal of low density bins, removing long edges, choice of bins-->

### Illustration on simulated data

### 2024-05-14 paper.tex
Plan for writing

The 2D hexagon grid is defined by the number of bins in each direction
\((b_1, b_2)\), as given by the centroids of each hexagon \((c_1, c_2)\)
and the lower left position where the grid starts at \((s_1, s_2)\),
which correspond to the lowest left centroid. The values of \(s_i\) need
to be below their respective minimum variable values, and could be a
full bin lower, to allow a buffer (\(q\)) corresponding to a full
hexagon width (\(a_1\)) and height (\(a_2\)) around the data. The values
of \(b_i\) are variables to be computed that define the reduction in
size of the data (\(n\) to \(m\)).

@@

\subsubsection{Obtaining bin
representations}\label{obtaining-bin-representations}

Let \(h : \mathbb{R}^{n\times k} \rightarrow \mathbb{R}^{m\times k}\) be
the function that maps a point in \kD{} space to it's bin representation
(\(C^{(k)}\)). This is the model points in \kD{} space.

@@

\subsubsection{Generating edges}\label{generating-edges}

Delaunay triangulation on bin representations generates edges by
specifying which bin representations should be connected to generate the
model in \kD{} space.

@@

The final step of the algorithm is \emph{lifting} the \kD{} model back
to \pD{}, completing the \emph{model-in-the-data-space} picture. For
this, define the set \(H^k\) to be the set of points that belong to
centroid \(k\). That is, \(H^k = \{ y \; | \; g(y) = (h_x^k, h_y^k)\}\).
We use the \pD{} Euclidean mean of the points in \(H^k\) to map the
centroid \((h_x^k, h_y^k)\) to a point in \pD{}. Let the \(i\)-th
component of the \pD{} mean be

@@

\subsection{What can be learned}\label{what-can-be-learned}

\begin{itemize}
\tightlist
\item
  Overview: Generate a form that maps the model, that is the interpoint
  distances. What is the model?
\item
  Notation
\item
  Create a representation of the model

  \begin{itemize}
  \tightlist
  \item
    using hex-binning in 2D,
  \item
    parameters,
  \item
    tuning,
  \item
    pre-processing
  \end{itemize}
\item
  How does this map to the representation in high-d

  \begin{itemize}
  \tightlist
  \item
    Centroids,
  \item
    Edges
  \end{itemize}
\item
  Measuring fit

  \begin{itemize}
  \tightlist
  \item
    Fitted values
  \item
    Error calculation
  \end{itemize}
\item
  What is learned about simulated examples

  \begin{itemize}
  \tightlist
  \item
    Interesting organisation of points in UMAP
  \item
  \end{itemize}
\end{itemize}

>>>>

The 2D hexagon grid is defined by the number of bins in each direction
\((b_1, b_2)\), giving total number of bins as \(b = b_1 \times b_2\),
and hexagon id, \(h = 1, \dots, b\). Each hexagon, \(H_h\) is uniquely
described by centroid, \$ C\_\{h\}\^{}\{(2)\} = (c\_\{h1\},
c\_\{h2\})\$. The lower left position where the grid starts at
\((s_1, s_2)\), which correspond to the lowest left centroid. The values
of \(s_i\) need to be below their respective minimum variable values,
and could be a full bin lower, to allow a buffer (\(q\)) corresponding
to a full hexagon width (\(a_1\)) and height (\(a_2\)) around the data.
The values of \(b_i\) are variables to be computed that define the
reduction in size of the data (\(n\) to \(m\)).

@@

\(\{ i \in H_h, h = 1, \dots, b, \text{ and } i = 1, \dots, n\}\)

\subsubsection{Summarization of the data in
2D}\label{summarization-of-the-data-in-2d}

Let \(h : \mathbb{R}^{n\times k} \rightarrow \mathbb{R}^{m\times k}\) be
the function that maps a point in \kD{} space to it's bin centroid
(\(C^{(k)}\)). This is the model points in \kD{} space.

@@

\subsubsection{Generating edges to indicate neighbours and removing long
edges}\label{generating-edges-to-indicate-neighbours-and-removing-long-edges}

Delaunay triangulation is used to connect neighbouring centroids, which
is needed to preserve neighbourhood information when the model is lifted
into \pD{}.

@@

The last step is to lift the \kD{} model into \pD{} by computing \pD{}
vectors that represent bin centroids. Define the set \(H^k\) to be the
set of points that belong to bin \(k\). We use the \pD{} Euclidean mean
of the points in \(H^k\) to map the centroid \((h_x^k, h_y^k)\) to a
point in \pD{}. Let the \(i\)-th component of the \pD{} mean be

@@

\subsection{Tuning}\label{tuning}

\subsection{Illustration on simulated
data}\label{illustration-on-simulated-data}
